Text Is All You Need

### 2023-05-05 docs/hes_final/article.qmd
Minor Grammar edits

We developed three approaches for extracting conflict signals from text embeddings. The first approach used supervised dimensionality reduction and a feed-forward neural network conditioned with country context characteristics. In the second approach we used a compact dataset of monthly embedding averages per country per month and applied statistical classification models such as K-Nearest Neighbors and XGBoost. In the third approach we stacked multiple article-level embeddings and applied Transformer-based models to extract a signal directly from the text corpus. See the @fig-ffnn, @fig-xgboost, and @fig-transformer for details of modeling approaches.

@@

Further investigation showed that while overall numeric metrics look similar, different models appeared to perform better in different regions of the world. While performance is often the main objective to optimize for when developing prediction models, our practical goal was to develop a model that can assist UNICEF in their decision making. With that in mind, cost of deployment is also an important factor for model evaluation. Average embeddings dataset and the corresponding statistical model fit in memory while working with article level embeddings in two other proposed methods requires additional data engineering steps to be processed in batches.

@@

If models like these were to become widely used, they may be vulnerable to adversarial attacks. Large volumes of fake news content could be used to overwhelm the true signal from news articles to obscure the intentions of malicious actors, or to cause panic. Finally, and perhaps most concerning, is the potential for conflict prediction models to deliver self-fulfilling prophesies. If policy makers become convinced of impending conflict in their own or neighboring countries because of the predictions of an AI model, they may take actions that make conflict more likely.

@@

Due to time and resource constraints, our work used only 20% of the daily embeddings available through  GDELT, and future research could focus on scaling the methods outline. We also believe that continued experimentation with data sampling and aggregation approaches, as well as model tuning, could yield models that outperform those outlined here without additional data.

@@

Building an application with a user interface could be a natural extension of this work. We envision a dashboard where a user can view a world map where countries are colored according to the estimated probability of escalating conflict. Countries with highest risk can be flagged or shown to a user in a separate view. In addition to the visual part of this application, model deployment for easy inference is an important project.

>>>>

We developed three approaches for extracting conflict signals from text embeddings. The first approach used supervised dimensionality reduction and a feed-forward neural network conditioned with country context characteristics. In the second approach, we used a compact dataset of monthly embedding averages per country per month and applied statistical classification models such as K-Nearest Neighbors and XGBoost. In the third approach we stacked multiple article-level embeddings and applied transformer-based models to extract a signal directly from the text corpus. See the @fig-ffnn, @fig-xgboost, and @fig-transformer for details of modeling approaches.

@@

Further investigation showed that while overall numeric metrics look similar, different models appeared to perform better in different regions of the world—see @fig-ffnn-test-accuracy, @fig-xgboost-test-accuracy, and @fig-transformer-test-accuracy. While performance is often the main objective to optimize for when developing prediction models, our practical goal was to develop a model that can assist UNICEF in their decision making. With that in mind, cost of deployment is also an important factor for model evaluation. Average embeddings dataset and the corresponding statistical model fit in memory while working with article level embeddings in two other proposed methods requires additional data engineering steps to be processed in batches.

@@

If models like these were to become widely used, they may be vulnerable to adversarial attacks. Large volumes of fake news content could be used to overwhelm the true signal from news articles to obscure the intentions of malicious actors, or to cause panic.

Finally, and perhaps most concerning, is the potential for conflict prediction models to deliver self-fulfilling prophesies. If policy makers become convinced of impending conflict in their own or neighboring countries because of the predictions of an AI model, they may take actions that make conflict more likely.

@@

Future research could focus on scaling the methods outline. Due to time and resource constraints, our work used only 20% of the daily embeddings available through  GDELT, and increasing the scale of training data could yield better results. We also believe that continued experimentation with data sampling and aggregation approaches, as well as model tuning, could produce models that outperform those outlined here without additional data.

@@

Building an application with a user interface could be a natural extension of this research. We envision a dashboard where a user can view a world map where countries are colored according to the estimated probability of escalating conflict. Countries with highest risk can be flagged or shown to a user in a separate view. In addition to the visual part of this application, model deployment for easy inference is an important project.
